## Introduction to dplyer

dplyr is an R package in the tidyverse, we can load the package using

```{r}
# install.packages("tidyverse")

library(dplyr)
library(ggplot2)
?dplyr

# preview dataset
head(starwars)


```

### filtering data
we can filter data using `filter()`, this allows us to subset observations (rows) based on their values (columns)

tips:

 * be sure u spell the column name correctly (and the value name if it's a categorical variable). remember R is case-sensitive
 * be sure to use `==` when comparing observations. (remember `=` is an assignment operator)
 * you can use >, <, >=, <=, and != to compare numeric or categorical variables (nominal variables are ranked alphabetically, while ordinal values have a built-in rank)

```{r}

filter(starwars, hair_color == "blond" | hair_color == "blonde")
filter(starwars, hair_color == "blond" & species == "Human" )
filter(starwars, species == "Human", homeworld == "Tatooine", skin_color == "fair")
# the bar (|) stands for 'or', 'and' is the and sign (&) or the comma (,)

filter(starwars, mass >= 75, mass <= 100, hair_color == "brown")
filter(starwars, name < "B")
filter(starwars, hair_color %in% c("blond", "brown", "blonde", "black"))

importantPlanets <- c("Tatooine", "Naboo")
filter(starwars, homeworld %in% importantPlanets)

# %in% is similar to 'or'

filteredData <- filter(starwars, species == "Human", homeworld == "Tatooine", skin_color == "fair")

#ranked data
head(diamonds) #displays first 6 rows
class(diamonds$cut) #gives u the specific types of data
summary(diamonds$cut) #gives u a count of each category or summary statistics if numeric
sumary(diamonds$carat)
head(diamonds$cut) #displays first 6 values (and levels)

filter(diamonds, cut > "Good")

#ordering categorical data
unique(starwars$eye_color)
factor(starwars$eye_color, c("red", "red, blue", "orange", " yellow", "green, yellow", "blue", "blue-gray", "pink", "hazel","brown", "dark", "gold", "black", "white", "unknown"), ordered = TRUE)

#find all characters that are shorter than 100 cm
filter(starwars, height < 100, species != "Droid")

#which characters were born between the years 100 and 200 (inclusive)
filter(starwars, birth_year >= 100, birth_year <= 200)

#which characters weigh over 100kg but are shorter than 185cm
filter(starwars, height < 185, mass > 100)

#missing values
NA > 5 #it doesn't know
is.na(10 == NA) #returns true or false

#which characters are missing a hair color
filter(starwars, is.na(hair_color))
```

### arranging data
`arrange()` reorders rows, it does not remove any rows. NA values are always at the end when you order by a column

```{r}
#lowest to highest
arrange(starwars, birth_year)

#highest to lowest
arrange(starwars, desc(birth_year))

#categorical is alphabetical
arrange(starwars, hair_color)

#multiple columns
arrange(starwars, hair_color, birth_year)

#arrange starwars characters to find the tallest characters and the shortest characters
arrange(starwars, height)

#how could u use arrange() to sort all missing variables to the start
arrange(starwars, !(is.na(starwars)))
arrange(starwars, desc(is.na(starwars)))

```

### we can select certain columns in the dataset

`select()` allows us to retain only certain variables 

```{r}

select(starwars, hair_color, skin_color, eye_color)
select(starwars, hair_color:eye_color) #returns every column between first:last

select(starwars, -(hair_color:eye_color)) #all except
select(starwars, -skincolor)

starwarsNoColor <- select(starwars, -(hair_color:eye_color))
ggplot(starwarsNoColor, aes(x = hair_color)) #error because we removed it

select(starwars, contains("color"))
select(starwars, ends_with("color"))
select(starwars, starts_with("color"))

```


### adding new columns
`mutate()`

```{r}

starwars_small <- select( starwars, height, mass, birth_year)
head(starwars_small)

mutate(starwars_small, 
       height_m = height/100, 
       bmi = mass/(height_m**2))

#to only keep new columns use transmute
transmute(starwars_small, 
       height_m = height/100, 
       bmi = mass/(height_m**2))

#using agregate functions
prop_mass <- mutate(starwars_small, proportional_mass = mass/sum(mass, na.rm = TRUE))
arrange(prop_mass, desc(propotional_mass))

```

### summarising and grouping data

`summarise()` collapses an entire column of data to a single value

```{r}

mutate(starwars, 
       mean_mass = mean(mass, na.rm = TRUE))

summarise(starwars, 
          mean_mass = mean(mass, na.rm = TRUE),
          mean_height = mean(height, na.rm = TRUE))

mean(starwars$mass, na.rm = TRUE)

speciesMasses <- summarise(group_by(starwars, species),
                           mean_mass = mean(mass, na.rm = TRUE),
                           count = n()) 
#n() returns the number of values in each species

arrange(speciesMasses, desc(mean_mass))

#the pipe operator: %>%
#function(x, y) is the same as x %>% function(y)
#when using dplyr functions, u generally always start with the dataset

speciesMasses <- starwars %>% 
  group_by(species) %>%
  summarise(mean_mass = mean(mass, na.rm = TRUE),
           count = n()) %>% 
  arrange(desc(mean_mass))


```

### sampling a designated number of rows
`sample_n()` allows us to sample a random number of rows from our dataset

```{r}

#10 random rows
starwars_10rows <- sample_n(starwars, 10)

starwars_10rows

#10% off rows, selected randomly
nrow(starwars) #total number of rows is 87
starwars_10percent <- sample_frac(starwars, 0.1)
starwars_10percent #9 rows is 10%



```

we can also take a "slice" of our dataset using `slice()` and its related set of functions

```{r}

slice(starwars, 50:60) #slice by index value (row number)

slice_head(starwars, 10) #first 10 rows by index
slice_tail(starwars, 10) #last 10 rows by index
slice_min(starwars, order_by = height, n = 10) #lowest 10 heights
slice_max(starwars, order_by = mass, n = 10) #largest 10 masses

```
