practice subsetting data
use a combination of filter, select, mutate, arrange, summarise, group_by, sample, and/or slice
create a visualization using ur new subset of data

```{r}
#download initial libraries
library(ggplot2)
library(dplyr)
show_data <- read.csv("data/tv_shows.csv")

head(show_data)

#first practice
cleanData <- select(show_data, Title:Rotten.Tomatoes) %>%
                filter(Age != "")

ggplot(data = cleanData, aes(x = factor(Age, c("all", "7+","13+","16+","18+")), fill = Age)) +
  geom_bar() +
  labs(x = "age ratings",
       y = "# of tv shows",
       title = "age ratings of popular tv shows",
       fill = "age ratings")
```


### filtering observations based on one or more variables
```{r}
mediocre_shows <- filter(cleanData, Rotten.Tomatoes <= "80/100", Rotten.Tomatoes != "100/100", Rotten.Tomatoes >= "60/100")

mediocre_shows #these are all of the middle-rated TV shows

oldies <- filter(cleanData, Year < 2000)
oldies #these are all the TV shows from the 1900s

new <- filter(cleanData, Year >= 2000)
new #these are all the TV shows from the 2000s

niche <- filter(cleanData, Rotten.Tomatoes > "80/100" | Rotten.Tomatoes == "100/100", Year > 2005, Title > "The", Title < "U")
niche #only rated 80 or higher, newer than 2005, titles starting with "the"

```

### creating a smaller dataset with a subset of variables

```{r}

miniData <- select(cleanData, contains("t") | starts_with("I") | starts_with("Y"))
head(miniData) #only includes the titles, years, and ratings

```

### adding 2 new columns using mutate()

```{r}

subbedTomatoes <- sub("/100","", miniData$Rotten.Tomatoes)

subbedIMDb <- sub("/10","", miniData$IMDb)

miniData <- filter(miniData, Year < 2000)

numberData <- mutate(miniData, 
                     numeric_rating = ((as.numeric(subbedTomatoes)) + (as.numeric(subbedIMDb) * 10)) / 2)

numberData <- mutate(numberData,
                     rating_dif = abs(as.numeric(subbedTomatoes) - (as.numeric(subbedIMDb)*10)))

numberData #gives numerical rating values to each title, includes year
```

### creating a data frame of grouped summaries on numeric variables by categorical variables

```{r}

summarise(numberData, 
          mean_rating = mean(numeric_rating, na.rm = TRUE),
          mean_dif = mean(rating_dif, na.rm = TRUE))

```

### creating new visualizations using an updated dataset

```{r}

ggplot(data = numberData, aes(x = numeric_rating)) +
  geom_histogram(fill = "pink", color = "black") +
  scale_x_continuous(breaks = seq(from = 20, to = 100, by = 10)) +
  labs(x = "average rating",
       y = "# of shows",
       title = "ratings of shows from before 2000")

```